### 2019-01-30 ic-journal/ccs.tex
Stage work

We assume a possibly infinite set of names $\mathscr{L} = \{a, b,
\ldots\}$ forming input and $\overline{\mbox{output}}$ actions, plus a special invisible
action $\tau \notin \mathscr{L}$, and a set of variables $A, B,
\ldots$ for defining recursive behaviours.
Given a deadlock $\nil$, the class of CCS processes is then inductively defined from $\nil$ by the operators
of prefixing, parallel composition, summation (binary choice), restriction, recursion and relabeling:
\begin{equation*}
\begin{array}{cccl}
\mu  & := & \tau\!\!\!\! & \midd \; a  \; \midd \;  \outC a  \\
P  & := & \nil\!\!\!\! & \midd \;  \mu . P \; \midd \;  P_1 |  P_2 \; \midd  \;
P_1 + P_2 \; \midd % \; \mu . P\; \midd  \; 
  (\res a\!)\, P  \;  \midd \;  A \; \midd \; \recu A  P
\; \midd \; P\; [r\!f]  % relabelling
\end{array}
\end{equation*}
\hl{We sometimes omit the trailing $\nil$, e.g., writing $a|b$ for $a.\nil |b .\nil$.}
The operational semantics of CCS is then given by means of
a Labeled Transition System (LTS), shown in Fig.~\ref{f:LTSCCS} as
some \hl{Structural Operational Semantics (SOS)}
rules (the symmetric version of the two rules for
parallel composition and the rule for sum are omitted).
A CCS expression uses only \emph{weakly-guarded sums} if all occurrences of
the sum operator are of the form $\mu_1.P_1 + \mu_2.P_2 + \ldots
+ \mu_n.P_n$, for some $n \geq 2$.
 The \emph{immediate derivatives} of a
process $P$ are the elements of the set $\{P' \st P \arr\mu P' \mbox{
  for some $\mu$}\}$.
 \hl{We use $\ell$ to range over
  visible actions (i.e.~inputs or outputs, excluding  $\tau$).}
\begin{figure*}[t]
\begin{center}
\vskip .1cm
 $\displaystyle{  \over  \mu.  P    \arr\mu
P } $  $ \hb$   
\hskip .5cm
 $\displaystyle{   P \arr\mu   P' \over   P + Q   \arr\mu
P'  } $  $ \hb$   
\hskip .5cm
 $\displaystyle{   P \arr\mu   P' \over   P | Q   \arr\mu
P' | Q } $  $ \hb$   
\hskip .3cm
  $\; \;$  $\displaystyle{ P \arr{ a}P' \hk \hk  Q
\arr{\outC a }Q'  \over     P|  Q \arr{ \tau} P'
|  Q'  }$ 
\\
\vspace{.2cm}
$\displaystyle{ P \arr{\mu}P' \over
 (\res a\!)\, P   \arr{\mu} (\res a\!)\, P'} $ $ \mu \neq a, \outC a$
$ \hb$
%
$\displaystyle{ P \sub {\recu A P} A \arr{\mu}P' \over
 \recu A P   \arr{ \mu} P'  } $
\hskip .5cm  
$\displaystyle{ P \arr{\mu} P' \over
 P \;[r\!f] \arr{r\!f(\mu)} P' \;[r\!f]} $ $\forall a.\, r\!f(\outC a) = \overline{r\!f(a)}$
$ \hb$ %  &
\end{center}
\caption{Structural Operational Semantics of CCS}
\label{f:LTSCCS}
\end{figure*}
Some standard notations for transitions: $\Arr\epsilon$ is the 
reflexive and transitive closure of $\arr\tau$, and 
$\Arr \mu $ is $\Arr\epsilon \arr\mu \Arr\epsilon$ (the
composition of the three relations).
Moreover,   
$ P \arcap \mu P'$ holds if $P \arr\mu P'$ or ($\mu =\tau$ and
$P=P'$); similarly 
$ P \Arcap \mu P'$ holds if $P \Arr\mu P'$ or ($\mu =\tau$ and
$P=P'$).
We write $P \:(\arr\mu)^n P'$ if $P$ can become $P'$ after performing
$n$ $\mu$-transitions. Finally, $P \arr\mu$ holds if there is $P'$
with $P \arr\mu P'$, and similarly for other forms of transitions.

\paragraph{Further notations}
\hl{Let} letters $\R$, $\S$ range over relations \hl{of CCS processes}.
We use infix notation for relations, e.g., 
$P \RR Q$ means that $(P,Q) \in \R$.
We use tilded letters \hl{like $\til P$} to denote tuples \hl{(of processes)} with
countably many elements; thus the tuple may also be infinite \hl{(??)}.
All notations are extended to tuples componentwise,
e.g., $\til P \RR \til Q$ means that $P_i \RR Q_i$, for each  
component $i$ of the tuples $\til P$ and $\til Q$.
And $\ct{\til P}$ is the process obtained by replacing each hole
$\holei i$ of the  context $\qct$ with $P_i$.  
We write $
\ctx \R$ for the closure of a relation under contexts. Thus $P\: \ctx \R\: Q$
means that there are context $\qct$ and tuples $\til P,\til Q$ with
$P =  \ct{\til P}, Q =  \ct{\til Q}$ and $\til P \RR \til Q$.
We use the symbol $\DSdefi$ for abbreviations. For instance, $P \DSdefi G $, where
$G$ is some expression, means that $P$ stands for the expression $G$.
If $\leq$ is a preorder, then  $\geq$  is its inverse (and
conversely).

>>>>

We assume a possibly infinite set of names $\mathscr{L} = \{a, b,
\ldots\}$ forming input and $\overline{\mbox{output}}$ actions, plus a special invisible
action $\tau \notin \mathscr{L}$, and a set of variables $A, B,
\ldots$ for defining recursive behaviours.
Given a deadlock $\nil$, the class of CCS processes is inductively defined from $\nil$ by the operators
of prefixing, parallel composition, summation (binary choice), restriction, recursion and relabeling:
\begin{equation*}
\begin{array}{cccl}
\mu  & := & \tau\!\!\!\! & \midd \; a  \; \midd \;  \outC a  \\
P  & := & \nil\!\!\!\! & \midd \;  \mu . P \; \midd \;  P_1 |  P_2 \; \midd  \;
P_1 + P_2 \; \midd % \; \mu . P\; \midd  \; 
  (\res a\!)\, P  \;  \midd \;  A \; \midd \; \recu A  P
\; \midd \; P\; [r\!f]  % relabelling
\end{array}
\end{equation*}
\hl{We sometimes omit the trailing $\nil$, e.g., writing $a|b$ for $a.\nil |b .\nil$.}
The operational semantics of CCS is then given by means of
a Labeled Transition System (LTS), shown in Fig.~\ref{f:LTSCCS} as
\hl{Structural Operational Semantics (SOS)}
rules (the symmetric version of the two rules for
parallel composition and the rule for sum are omitted).
A CCS expression uses only \emph{weakly\hl{ }guarded sums} if all occurrences of
the sum operator are of the form $\mu_1.P_1 + \mu_2.P_2 + \ldots
+ \mu_n.P_n$, for some $n \geq 2$.
 The \emph{immediate derivatives} of a
process $P$ are the elements of the set $\{P' \st P \arr\mu P' \mbox{
  for some $\mu$}\}$.
 \hl{We use $\ell$ to range over
  visible actions (i.e.~inputs or outputs, excluding  $\tau$).}
\begin{figure*}[t]
\begin{center}
\vskip .1cm
 $\displaystyle{  \over  \mu.  P    \arr\mu
P } $  $ \hb$   
\hskip .5cm
 $\displaystyle{   P \arr\mu   P' \over   P + Q   \arr\mu
P'  } $  $ \hb$   
\hskip .5cm
 $\displaystyle{   P \arr\mu   P' \over   P | Q   \arr\mu
P' | Q } $  $ \hb$   
\hskip .3cm
  $\; \;$  $\displaystyle{ P \arr{ a}P' \hk \hk  Q
\arr{\outC a }Q'  \over     P|  Q \arr{ \tau} P'
|  Q'  }$ 
\\
\vspace{.2cm}
$\displaystyle{ P \arr{\mu}P' \over
 (\res a\!)\, P   \arr{\mu} (\res a\!)\, P'} $ $ \mu \neq a, \outC a$
$ \hb$
%
$\displaystyle{ P \sub {\recu A P} A \arr{\mu}P' \over
 \recu A P   \arr{ \mu} P'  } $
\hskip .5cm  
$\displaystyle{ P \arr{\mu} P' \over
 P \;[r\!f] \arr{r\!f(\mu)} P' \;[r\!f]} $ $\forall a.\, r\!f(\outC a) = \overline{r\!f(a)}$
$ \hb$ %  &
\end{center}
\caption{Structural Operational Semantics of CCS}
\label{f:LTSCCS}
\end{figure*}
Some standard notations for transitions: $\Arr\epsilon$ is the 
reflexive and transitive closure of $\arr\tau$, and 
$\Arr \mu $ is $\Arr\epsilon \arr\mu \Arr\epsilon$ (the
composition of the three relations).
Moreover,   
$ P \arcap \mu P'$ holds if $P \arr\mu P'$ or ($\mu =\tau$ and
$P=P'$); similarly 
$ P \Arcap \mu P'$ holds if $P \Arr\mu P'$ or ($\mu =\tau$ and
$P=P'$).
We write $P \:(\arr\mu)^n P'$ if $P$ can become $P'$ after performing
$n$ $\mu$-transitions. Finally, $P \arr\mu$ holds if there is $P'$
with $P \arr\mu P'$, and similarly for other forms of transitions.

\paragraph{Further notations}
\hl{Let variables} $\R$, $\S$ range over binary relations%
\footnote{We use infix notation for relations, e.g., 
$P \RR Q$ means that $(P,Q) \in \R$.} of CCS processes.
We use tilded letters \hl{like $\til P$} to denote tuples \hl{of processes} with
countably many elements; (Thus the tuple may also be infinite.)
All notations are extended to tuples componentwise,
e.g., $\til P \RR \til Q$ means that $P_i \RR Q_i$, for each  
component $i$ of the tuples $\til P$ and $\til Q$.
And $\ct{\til P}$ is the process obtained by replacing each hole
$\holei i$ of the  context $\qct$ with $P_i$.  
We write $
\ctx \R$ for the closure of a relation under contexts. Thus $P\: \ctx \R\: Q$
means that there are context $\qct$ and tuples $\til P,\til Q$ with
$P =  \ct{\til P}, Q =  \ct{\til Q}$ and $\til P \RR \til Q$.
We use the symbol $\DSdefi$ for abbreviations. For instance, $P \DSdefi G $, where
$G$ is some expression, means that $P$ stands for the expression $G$.
If $\leq$ is a preorder, then  $\geq$  is its inverse (and
conversely).

### 2019-01-30 ic-journal/contraction.tex
Stage work

\begin{definition}%[bisimulation contraction]
\label{d:BisCon}
A process relation ${\R}$ 
 is a \textbf{(bisimulation) contraction} if, whenever
 $P\RR Q$, %for all $\mu$

\begin{enumerate}
\item   $P \arr\mu P'$ implies there is $Q'$ such that $Q \arcap \mu
  Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$   implies there is $P'$ such that $P \Arcap \mu
 P'$ and $P' \wb Q'$\enspace.
\end{enumerate}
\emph{Bisimilarity contraction}, written as $P \mcontrBIS Q$ ($P$
\textbf{contracts to} $Q$), if $P\ \R\ Q$ for some contraction $\R$.
\end{definition}
% We say that $P$  {\emph{bisimilarly  contracts} $Q$, written
% $Q \mcontrBIS Q$.
% We write $\mexpaBIS$ for the inverse of $\mcontrBIS$.

@@

Bisimilarity  contraction is coarser than 
 the \emph{expansion relation} 
$\expa$ \cite{arun1992efficiency,sangiorgi2015equations}.
This is a
preorder widely used in proof techniques for bisimilarity and that 
intuitively refines bisimilarity by 
 formalising the idea of \emph{efficiency} between processes.
Clause (1) is the same in the two
preorders. But in clause (2) expansion uses 
$P \Arr \mu P'$, rather than $P \Arcap \mu P'$; 
 moreover with
contraction the final derivatives are simply required to be bisimilar.
An expansion 
$P \expa Q$
tells us  that $Q$ is always at least as efficient as $P$, whereas  the
 contraction $P \mcontrBIS Q$  just says that $Q$ has the  possibility of
being at least as efficient as $P$.

@@

The substitutivity proofs for for expansion and bisimilarity carry
over to contraction. Like (weak) bisimilarity and expansion, contraction is
preserved by all operators but (direct) sum.

% next file: unique.tex

>>>>

\begin{definition}%[bisimulation contraction]
\label{d:BisCon}
A process relation ${\R}$ 
 is a \textbf{contraction} if, whenever
 $P\RR Q$, for all $\mu$
\begin{enumerate}
\item $P \arr\mu P'$ implies there is $Q'$ such that $Q \arcap \mu
  Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$   implies there is $P'$ such that $P \Arcap \mu
 P'$ and $P' \wb Q'$\enspace.
\end{enumerate}
\emph{Bisimilarity contraction}, written as $P \mcontrBIS Q$ ($P$
\textbf{contracts to} $Q$), if $P\ \R\ Q$ for some contraction $\R$.
\end{definition}

@@

Bisimilarity contraction is coarser than 
 the \emph{expansion relation} $\expa$.
This is a preorder widely used in proof techniques for bisimilarity and that 
intuitively refines bisimilarity by 
 formalising the idea of \emph{efficiency} between processes.
Clause (1) is the same in the two
preorders. But in clause (2) expansion uses 
$P \Arr \mu P'$, rather than $P \Arcap \mu P'$; 
 moreover with
contraction the final derivatives are simply required to be bisimilar.
An expansion 
$P \expa Q$
tells us  that $Q$ is always at least as efficient as $P$, whereas  the
 contraction $P \mcontrBIS Q$  just says that $Q$ has the  possibility of
being at least as efficient as $P$.

@@

The substitutivity proofs for for expansion and bisimilarity carry
over to contraction. Like (weak) bisimilarity and expansion, contraction is
preserved by all operators but (direct) sums.

% next file: equa.tex

### 2019-01-30 ic-journal/equa.tex
Stage work

\section{Equations and contractions}
\label{s:eq}

@@

Uniqueness of  solutions of equations \cite{Mil89} intuitively says that if  a context $\qct$ obeys
certain  conditions, 
then all processes $P$  that satisfy the equation $ P \wb \ct P$ are
bisimilar with each other.
We need variables to write equations. We use capital letters $X,Y,Z$
for these variables and call them \emph{\behav\  variables}.
The body of an equation is a CCS expression
possibly containing \behav\  variables. Thus such expressions, ranged
over by $E$, live in the CCS grammar extended with \behav\  variables.
% , which we  call 
%  \emph{extended CCS}.

\begin{definition}[\hl{System of equations}]
Assume that, for each $i$ of 
 a countable indexing set $I$, we have variables $X_i$, and expressions
$E_i$ possibly containing  such variables. 
Then 
$\{  X_i = E_i\}_{i\in I}$
is 
  a \emph{system of equations}. (There is one equation for each variable $X_i$.)
\end{definition}

@@

\begin{theorem}[Unique solution of equations \cite{Mil89}]
\label{t:Mil89}
A system of guarded and sequential equations (without direct sums)
$\{  X_i = E_i\}_{i\in I}$ has \hl{an} unique solution for $\wb$.
\end{theorem}

% The proof exploits an invariance property on immediate transitions for
% guarded and sequential expressions, and then extracts a bisimulation
% (up-to bisimilarity) out
% of the solutions of the system.  
To see the  need of the sequentiality  condition, consider
 the equation (from \cite{Mil89}) $X = \res a (a. X | \outC a)$
where $X$ is guarded but not sequential. Any process that does not use $a$ is a solution.

% next file: contraction.tex

>>>>

\section{Systems of equations and contractions}
\label{s:eq}

@@

Uniqueness of  solutions of equations \cite{Mil89} intuitively says that if  a context $\qct$ obeys
certain  conditions, 
then all processes $P$  that satisfy the equation $ P \wb \ct P$ are
bisimilar with each other.
We need variables to write equations. We use capital letters $X,Y,Z$
for these variables and call them \emph{\behav\  variables}.
The body of an equation is a CCS expression
possibly containing \behav\  variables. Thus such expressions, ranged
over $E$, live in the CCS grammar extended with \behav\ variables
, which we  call \emph{extended CCS}.

\begin{definition}[\hl{System of equations}]
Assume that, for each $i$ of 
 a countable indexing set $I$, we have variables $X_i$, and expressions
$E_i$ possibly containing  such variables. 
Then $\{X_i = E_i\}_{i\in I}$ is 
a \emph{system of equations}. (There is one equation for each variable $X_i$.)
\end{definition}

@@

\begin{theorem}[Unique solution of equations \cite{Mil89}]
\label{t:Mil89}
A system of guarded and sequential equations (without direct sums)
$\{X_i = E_i\}_{i\in I}$ has \hl{an} unique solution for $\wb$.
\end{theorem}

The proof exploits an invariance property on immediate transitions for
guarded and sequential expressions, and then extracts a bisimulation
(up-to bisimilarity) out
of the solutions of the system.
To see the  need of the sequentiality condition, consider
 the equation (from \cite{Mil89}) $X = \res a (a. X | \outC a)$
where $X$ is guarded but not sequential. Any process that does not use $a$ is a solution.

% next file: unique.tex

### 2019-01-30 ic-journal/expansion.tex
Stage work

The equivalences we consider here are mainly \emph{weak} ones, in that they
abstract from the number of internal steps being performed:
\begin{definition}%[bisimilarity]
\label{d:wb}
A process relation ${\R}$ is a \textbf{bisimulation} if, whenever
 $P\RR Q$, %for all $\mu$ 
we have:
\begin{enumerate}
\item $P \arr\mu P'$ implies that there is $Q'$ such that $Q \Arcap \mu Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$,implies that there is $P'$ such that $P \Arcap
  \mu P'$ and $P' \RR Q'$\enspace.
% (the converse of (1) on the actions from $Q$)
\end{enumerate}  
 $P$ and $Q$ are \textbf{bisimilar},
written as $P \wb Q$, if $P \RR Q$ for some bisimulation $\R$.
\end{definition}

We sometimes call bisimilarity the \emph{weak} one, to
distinguish it from \emph{strong} bisimilarity ($\sim$),
obtained by replacing in the above definition   the weak answer $
Q\Arcap\mu Q'$ with the strong  $Q \arr \mu Q'$.
Weak bisimilarity is not preserved by the sum operator (except for
guarded sums). For this, Milner introduced \emph{observational congruence}, also called \emph{rooted
  bisimilarity} \cite{Gorrieri:2015jt,Sangiorgi:2011ut}:
\begin{definition}%[rooted bisimilarity]
\label{d:rootedBisimilarity}
Two processes $P$ and $Q$ are \textbf{rooted bisimilar}, written as $P
\rapprox Q$, if we have:
\begin{enumerate}
 \item  $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \wb Q'$;
 \item  $Q \arr\mu Q'$ implies that there is $P'$ such that $P
   \Arr\mu P'$ and $P' \wb Q'$\enspace.
%the converse of (1) on the actions from $Q$.
\end{enumerate}
\end{definition}
% Besides reducing the rooted bisimiarity of two processes to
% the bisimilarities of their first-step transition ends, this definition also brings a proof technique for proving the
% rooted bisimiarity by constructing a bisimulation:
% \begin{lemma}{(Rooted bisimilarity by constructing a bisimulation)}
% \label{l:obsCongrByWeakBisim}
% Given a (weak) bisimulation $\RR$, if two processes $P$ and $Q$
% satisfies the following properties:
% \begin{enumerate}
%  \item  $P \arr\mu P'$ implies that there is $Q'$ such that $Q
%    \Arr\mu Q'$ and $P' \RR Q'$;
% \item the converse of (1) on the actions from $Q$.
% \end{enumerate}
% then $P$ and $Q$ are rooted bisimilar, i.e.~$P \approx^c Q$.
% \end{lemma}

\begin{theorem}
\label{t:rapproxCongruence}
$\rapprox$ is a congruence in CCS, and it is the
coarsest congruence contained in $\approx$.
\end{theorem}

% \finish{Remove expansion here? --Chun}
% The bisimulation proof method can be enhanced by means of \emph{up-to
%   techniques}. One of the most useful auxiliary relations in up-to
% techniques  is the \emph{expansion} relation  $\expa$ \cite{SaMi92}. This is an asymmetric version
% of $\wb$ where $P \expa Q$  means that 
%  $P \wb Q$,
% but also that $Q$  achieves  the same as  $P$ 
% with  no more work, i.e.~with no more $\tau$ actions.
% Intuitively, if $ P \expa Q$, we can think of $Q$ as being 
% at least as fast as $P$
% or, more generally, we can think that $P$  uses at least as many resources as $Q$.

% \begin{definition}[expansion]
% \label{d:expa}
% A process relation ${\R}$ 
%  is an {\em  expansion} if, whenever
% we have $P\RR Q$, %for all $\mu$ 
% \begin{enumerate}
% \item   $P \arr\mu P'$ implies that there is $Q'$ with $Q \arcap \mu
%   Q'$
%  and $P' \RR Q'$;
% \item 
%     $Q \arr\mu Q'$   implies that there is $P'$ with $P \Arr \mu
%  P'$ and $P' 
% \RR Q'$.
% \end{enumerate}
%  $P$  {\em expands} $Q$, written
% $P  \expa Q$, 
% if $P \RR Q$,  for some expansion $\R$. 
% \end{definition}

% As bisimilarity, expansion is preserved by all operators but sum.
% An example of up-to technique involving expansion is

% next file: equa.tex

>>>>

The equivalences we consider here are mainly \emph{weak} ones, in that they
abstract from the number of internal steps being performed:
\begin{definition}%[bisimilarity]
\label{d:wb}
A process relation ${\R}$ is a \textbf{bisimulation} if, whenever
 $P\RR Q$, \hl{for all $\mu\in \mathscr{L}\cup\{\tau\}$} we have:
\begin{enumerate}
\item $P \arr\mu P'$ implies that there is $Q'$ such that $Q \Arcap \mu Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$,implies that there is $P'$ such that $P \Arcap
  \mu P'$ and $P' \RR Q'$\enspace.
\end{enumerate}
 $P$ and $Q$ are \textbf{bisimilar},
written as $P \wb Q$, if $P \RR Q$ for some bisimulation $\R$.
\end{definition}

We sometimes call bisimilarity the \emph{weak} one, to
distinguish it from \emph{strong} bisimilarity ($\sim$),
obtained by replacing in the above definition   the weak answer $
Q\Arcap\mu Q'$ with the strong  $Q \arr \mu Q'$.
Weak bisimilarity is not preserved by the sum operator (except for
guarded sums). For this, Milner introduced \emph{observational congruence}, also called \emph{rooted
  bisimilarity} \cite{Gorrieri:2015jt,Sangiorgi:2011ut}:
\begin{definition}%[rooted bisimilarity]
\label{d:rootedBisimilarity}
Two processes $P$ and $Q$ are \textbf{rooted bisimilar}, written as $P
\rapprox Q$, iff \hl{for all $\mu\in \mathscr{L}\cup\{\tau\}$}
\begin{enumerate}
 \item  $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \wb Q'$;
 \item  $Q \arr\mu Q'$ implies that there is $P'$ such that $P
   \Arr\mu P'$ and $P' \wb Q'$\enspace.
\end{enumerate}
\end{definition}
% Besides reducing the rooted bisimiarity of two processes to
% the bisimilarities of their first-step transition ends, this definition also brings a proof technique for proving the
% rooted bisimiarity by constructing a bisimulation:
% \begin{lemma}{(Rooted bisimilarity by constructing a bisimulation)}
% \label{l:obsCongrByWeakBisim}
% Given a (weak) bisimulation $\RR$, if two processes $P$ and $Q$
% satisfies the following properties:
% \begin{enumerate}
%  \item  $P \arr\mu P'$ implies that there is $Q'$ such that $Q
%    \Arr\mu Q'$ and $P' \RR Q'$;
% \item the converse of (1) on the actions from $Q$.
% \end{enumerate}
% then $P$ and $Q$ are rooted bisimilar, i.e.~$P \approx^c Q$.
% \end{lemma}

\begin{theorem}
\label{t:rapproxCongruence}
$\rapprox$ is a congruence in CCS, and it is the
coarsest congruence contained in $\approx$ \cite{van2005characterisation}.
\end{theorem}

\subsection{Expansions}
\label{s:expa}

\hl{The bisimulation proof method can be enhanced by means of \emph{up-to
techniques}. One of the most useful auxiliary relations in up-to
techniques is the \emph{expansion} relation} $\expa$ \cite{arun1992efficiency,sangiorgi2015equations}.
This is an asymmetric version
of $\wb$ where $P \expa Q$ means that $P \wb Q$,
but also that $Q$ achieves the same as $P$
with no more work, i.e.~with no more $\tau$ actions.
Intuitively, if $P \expa Q$, we can think of $Q$ as being
at least as fast as $P$
or, more generally, we can think that $P$ uses at least as many resources as $Q$.
\begin{definition}%[expansion]
\label{d:expa}
A process relation ${\R}$
  is an \textbf{expansion} if, whenever
we have $P\RR Q$, for all $\mu$
 \begin{enumerate}
 \item   $P \arr\mu P'$ implies that there is $Q'$ with $Q \arcap \mu
   Q'$
  and $P' \RR Q'$;
 \item
     $Q \arr\mu Q'$   implies that there is $P'$ with $P \Arr \mu
  P'$ and $P'
 \RR Q'$.
 \end{enumerate}
  $P$  {\em expands} $Q$, written as
 $P  \expa Q$,
 if $P \RR Q$ for some expansion $\R$.
 \end{definition}

Same as bisimilarity, the expansion preorder is preserved by all operators but (direct) sums.

% next file: contraction.tex

### 2019-01-30 ic-journal/intro.tex
Stage work

The main goal of the work described in this paper is a rather
comprehensive formalisation of the core theory of CCS in the HOL
theorem prover (HOL4),  with a focus on the theory of unique solutions
of \hl{equations and} contractions.
The formalisation, however, is not confined to the theory of unique
solutions of equations \hl{(and contractions)}, but embraces a
significant portion the theory of CCS \cite{Mil89}, mostly because the
theory of unique solutions relies on a large number of more fundamental results.
Indeed the formalisation encompasses the basic properties of strong and weak
bisimilarity (e.g. the fixed-point and substitutivity properties), the
basic properties of
rooted bisimilarity (the congruence induced by weak
bisimilarity, also called observation congruence), and
their algebraic laws. Further extensions (beyond Nesi
\cite{Nesi:1992ve}) include four versions of ``bisimulation up to''
techniques (e.g., bisimulation up-to bisimilarity) \cite{Mil89,sangiorgi1992problem}, and the
expansion and contraction preorder (two
efficiency-like refinements of weak bisimilarity). Concerning rooted bisimilarity, the formalisation
includes Hennessy Lemma and Deng Lemma (Lemma 4.1 and 4.2 of
\cite{Gorrieri:2015jt}),
 and two long proofs saying \hl{that} the rooted bisimilarity is the coarsest (largest)
 congruence contained in (weak) bisimilarity: one following Milner's
 book \cite{Mil89}, with the hypothesis that no processes can use up
 all labels;
the other without such hypothesis, essentially formalising van Glabbeek's paper \cite{van2005characterisation}.
Similar theorems are proved for the rooted contraction preorder.
In this respect, the work is an extensive experiment with the use of the HOL theorem prover with its
recent developments, including a package for coinductive definitions.

@@

Another \hl{benefit} of the formalisation is 
that we can take advantage of results about different 
equivalences and preorders that share similar proof structures. 
Examples are: the results that rooted bisimilarity and rooted contraction are,
respectively, the coarsest congruence contained in weak bisimilarity 
and the coarsest precongruence contained in the contraction preorder; 
the result about unique solution of equations for weak bisimilarity that uses the
contraction preorder as an auxiliary relation, and other unique solution results (e.g., 
the one for rooted in which
the auxiliary relation is rooted contraction); various forms of enhancements of the bisimulation
proof method (the `up-to' techniques).
In these cases, moving between proofs there are only a few places in
which the HOL proof scripts have to be modified.
Then the successful termination of the proof gives us a guarantee that the proof is
complete and trustworthy, \hl{eliminating} the risks 
of overlooking or missing details as in \hl{\emph{paper-and-pencil} proofs}.

@@

Section~\ref{ss:ccs} presents basic background materials on CCS,
including its syntax, operational semantics, bisimilarity and rooted
bisimilarity.
%
Section~\ref{s:eq} discusses equations and contractions.
%
Section~\ref{ss:new} presents rooted contraction and the related
unique-solution result for rooted bisimilarity.
%
Section~\ref{s:for} highlights our formalisation in HOL4.
%
Finally,  Section~\ref{s:rel} and \ref{s:concl} discuss related work,
conclusions, and a few directions for future work.

>>>>

The main goal of the work described in this paper is a rather
comprehensive formalisation of the core theory of CCS in the HOL
theorem prover (HOL4),  with a focus on the theory of unique solutions
of \hl{equations and} contractions.
The formalisation, however, is not confined to the theory of unique
solutions of equations \hl{(and contractions)}, but embraces a
significant portion the theory of CCS \cite{Mil89}, mostly because the
theory of unique solutions relies on a large number of more fundamental results.
Indeed the formalisation encompasses the basic properties of strong and weak
bisimilarity (e.g.~the fixed-point and substitutivity properties), the
basic properties of
rooted bisimilarity (the congruence induced by weak
bisimilarity, also called observation congruence), and
their algebraic laws. Further extensions (beyond Nesi
\cite{Nesi:1992ve}) include four versions of ``bisimulation up to''
techniques (e.g.~bisimulation up-to bisimilarity) \cite{Mil89,sangiorgi1992problem}, and the
expansion and contraction preorder (two
efficiency-like refinements of weak bisimilarity). Concerning rooted bisimilarity, the formalisation
includes Hennessy Lemma and Deng Lemma (Lemma 4.1 and 4.2 of
\cite{Gorrieri:2015jt}),
 and two long proofs saying \hl{that} the rooted bisimilarity is the coarsest (largest)
 congruence contained in (weak) bisimilarity: one following Milner's
 book \cite{Mil89}, with the hypothesis that no processes can use up
 all labels;
the other without such hypothesis, essentially formalising van Glabbeek's paper \cite{van2005characterisation}.
Similar theorems are proved for the rooted contraction preorder.
In this respect, the work is an extensive experiment with the use of the HOL theorem prover with its
recent developments, including a package for coinductive definitions.

@@

Another \hl{benefit} of the formalisation is 
that we can take advantage of results about different 
equivalences and preorders that share similar proof structures. 
Examples are: the results that rooted bisimilarity and rooted contraction are,
respectively, the coarsest congruence contained in weak bisimilarity 
and the coarsest precongruence contained in the contraction preorder; 
the result about unique solution of equations for weak bisimilarity that uses the
contraction preorder as an auxiliary relation, and other unique solution results (e.g.~%
the one for rooted in which
the auxiliary relation is rooted contraction); various forms of enhancements of the bisimulation
proof method (the `up-to' techniques).
In these cases, moving between proofs there are only a few places in
which the HOL proof scripts have to be modified.
Then the successful termination of the proof gives us a guarantee that the proof is
complete and trustworthy, \hl{eliminating} the risks 
of overlooking or missing details as in \hl{\emph{paper-and-pencil} proofs}.

@@

Section~\ref{ss:ccs} presents basic background materials on CCS,
including its syntax, operational semantics, bisimilarity and rooted
bisimilarity.
Section~\ref{s:eq} discusses equations and contractions.
Section~\ref{ss:new} presents rooted contraction and the related
unique-solution result for rooted bisimilarity.
Section~\ref{s:for} highlights our formalisation in HOL4.
Finally,  Section~\ref{s:rel} and \ref{s:concl} discuss related work,
conclusions, and a few directions for future work.
